---
title: "SpecData(Pollution)"
author: "Serge Ballif"
date: "May 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 2 R programming

```{r}
setwd("~/datasciencecoursera/Coursera")

pollutantmean <- function(directory, pollutant = "sulfate", id = 1:332) {

  # Set working directory
  if(grep("specdata", directory) == 1) {
    directory <- ("./specdata/")
  }

  # Locate files in specdata folder
  myfiles <- as.character( list.files(directory) )
  myfilepaths <- paste(directory, myfiles, sep="")
  
  # Initialize a vector to hold the pollutant data
  alldata <- c()
  
  # Loop through id files and collect the data values that are !na.
  for(index in id) {
    new <- read.csv(myfilepaths[index], header=T, sep=",")
    polcol <- new[,pollutant] # just pollutant column
    clean <- na.omit(polcol) # remove na
    alldata <- c(alldata, clean) # concatenate
  }
  
  # Calculate the mean and round to 3 decimal places.
  round(mean(alldata),digits=3)
}
```


```{r}
setwd("~/datasciencecoursera/Coursera")

complete <- function(directory, id = 1:332) {
  
  # Set working directory
  if(grep("specdata", directory) == 1) {
    directory <- ("./specdata/")
  }
  
  # Locate files in specdata folder
  myfiles <- as.character( list.files(directory) )
  myfilepaths <- paste(directory, myfiles, sep="")
  
  # Initialize a vector to hold the count of complete rows
  nobs <- c()
  
  # Loop through id files and count the complete rows.
  for(index in id) {
    new <- read.csv(myfilepaths[index], header=T, sep=",")
    nobs<- c(nobs,nrow(new[complete.cases(new),]))
  }
  
  # Create the data frame
  df=data.frame(id,nobs)
  df
}



```{r}
setwd("~/datasciencecoursera/Coursera")

source("complete.R")

corr <- function(directory, threshold = 0) {
  
  # Set working directory
  if(grep("specdata", directory) == 1) {
    directory <- ("./specdata/")
  }
  
  # Locate files in specdata folder
  myfiles <- as.character( list.files(directory) )
  myfilepaths <- paste(directory, myfiles, sep="")
  
  # call complete function to see which pass the threshold
  frame<-complete(directory)
  
  # Get data sets that pass the threshold
  pass<-subset(frame, nobs>threshold)
  id=pass[,1]
  
  # Define empty vector to store correlations
  corrvec<-c()
  
  # Loop through id files and collect correlations.
  for(index in id) {
    new <- read.csv(myfilepaths[index], header=T, sep=",")
    #newer <- na.omit(new) # remove na didn't seem to work. Hmm.
    #newcor=cor(new[,1],new[,2])
    newer <- new[complete.cases(new),] #remove na values
    
    # Compute correlation to 5 decimals
    newcor=round(cor(newer[,"sulfate"],newer[,"nitrate"]),5) 
    corrvec<-c(corrvec,newcor)
  }
  
  # Print
  corrvec
}

```

Alternatives that I came across
```{r}
corr <- function(directory, threshold = 0) {
  ## 'directory' is a character vector of length 1 indicating
  ## the location of the CSV files
  
  ## 'threshold' is a numeric vector of length 1 indicating the
  ## number of completely observed observations (on all
  ## variables) required to compute the correlation between
  ## nitrate and sulfate; the default is 0
  
  ## Return a numeric vector of correlations
  
  tcorr <- function(fname) {
    data <- read.csv(file.path(directory, fname))
    nobs <- sum(complete.cases(data))
    if (nobs > threshold) {
      return (cor(data$nitrate, data$sulfate, use="complete.obs"))
    }
  }
  tcorrs <- sapply(list.files(directory), tcorr) #get all correlations + NULLs
  tcorrs <- unlist(tcorrs[!sapply(tcorrs, is.null)]) #remove NULLs
  return (tcorrs)
}

corr <- function(directory, threshold = 0) {
  ## 'directory' is a character vector of length 1 indicating
  ## the location of the CSV files
  
  ## 'threshold' is a numeric vector of length 1 indicating the
  ## number of completely observed observations (on all
  ## variables) required to compute the correlation between
  ## nitrate and sulfate; the default is 0
  
  ## Return a numeric vector of correlations
  result <- c()
  all_files <- list.files(path=directory, full.names=T)
  
  for ( file_name in all_files){
    
    #path_to_file <- file.path(directory, paste(file_name,".csv", sep=""), fsep=.Platform$file.sep )
    #		print(path_to_file)
    #reads the data from the file
    #rm(airData)
    airData <- read.csv(file_name, header=T)
    num_complete_cases <- dim(airData[complete.cases(airData),])[1]
    if (num_complete_cases > threshold){
      
      correlation <- cor(airData$sulfate , airData$nitrate, use="complete.obs")
      result <- c(result, correlation)
    }
    
  }
  
  return(result)
  
  
}


```
